---
output: html_document
editor_options:
    chunk_output_type: console
---

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
library(patchwork)
options(digits = 3)
```

# Time Series Regression with Correlated Errors

The goal of this lecture is to bring together knowledge about regression, time series trends, and autocorrelation with the goal of obtaining correct inference in regression problems involving time series. 
You should become capable of suggesting a model that would meet the goal of data analysis (testing certain relationships) while satisfying assumptions (e.g., uncorrelatedness of residuals) and minimizing the risk of spurious results.

**Objectives**

1. Model autocorrelated residuals (in other words, incorporate or account for autocorrelation) so the final residuals satisfy modeling assumptions.
1. Apply and interpret Granger causality that is based on time series predictability.

**Reading materials**

* Chapters 3.8, 5.5, and 5.6 in @Shumway:Stoffer:2017
* Chapter 3 in @Kirchgassner:Wolters:2007 on Granger causality


## Introduction

Here we explore a regression model with autocorrelated residuals. We already know how to

* forecast, construct prediction intervals, and test hypotheses about regression coefficients when the residuals satisfy the ordinary least squares (OLS) assumptions, i.e.,  the residuals are white noise and have a joint normal distribution $N(0, \sigma^{2})$;
* model serial dependence (i.e., autocorrelation) in stationary univariate time series, such as using ARMA models.

Here we bring these two skills together.

If the residuals do not satisfy the independence assumption, we can sometimes describe them in terms of a specific correlation model involving one or more new parameters and some 'new residuals.' 
These new residuals essentially satisfy the OLS assumptions and can replace the original correlated residuals (which we will sometimes call the 'old' residuals) in the model.

When modeling time series, we often want to find so-called *leading indicators*, i.e., exogenous variables whose lagged values can be used for predicting the response, so for the forecasts we do not need to predict $X$-variables that are employed in a model. 
Hence, our models may include lagged versions of both the response and predictor variables (a general form of such a model was given in @eq-fmult).


## Linear regression with ARMA errors

We begin with the simplest case, a *constant mean model*, where the residuals are serially correlated and follow an AR(1) model, that is
$$
Y_{t} = \beta_{0} + \epsilon_{t},
$${#eq-constmean}
where $\epsilon_{t} \sim$ AR(1), i.e.,
$$
\epsilon_{t} = \phi \epsilon_{t - 1} + a_{t}.
$${#eq-ar1resid}

Here $\phi$ is a real number satisfying $0 < |\phi| < 1$; $a_{t}$ is white noise with zero mean and the variance $\nu^{2}$, i.e., $a_{t} \sim \mathrm{WN}(0,\nu^2)$.
We also assume that $\mathrm{cov}(a_{t}, \epsilon_{s}) = 0$ for all $s < t$ (i.e., that the residuals $\epsilon_s$ are not correlated with the future white noise $a_t$).

@eq-ar1resid allows us to express @eq-constmean as
$$
Y_{t} = \beta_{0} + \phi \epsilon_{t - 1} + a_{t}.
$${#eq-modelwithAR1}

The advantage of this representation is that the new residuals $a_{t}$ satisfy the OLS assumptions. 
In particular, since $a_{t}$ is white noise, $a_{t}$ is homoskedastic and uncorrelated.

We shall also assume that $a_{t}$ are normally distributed (for justification of the construction of confidence intervals and prediction intervals). 
However, even if $a_{t}$ are not normal, $a_{t}$ are uncorrelated, which is a big improvement over the serially correlated $\epsilon_{t}$.

Our goal is to remove the $\epsilon_{t}$ entirely from the constant mean model and replace them with $a_{t}$ acting as new residuals. 
This can be done as follows. 
First, write @eq-constmean for $t-1$ and multiply both sides by $\phi$,
$$
\phi Y_{t -1} = \phi \beta_{0} + \phi \epsilon_{t - 1}.
$${#eq-constmean1}

Taking the difference (@eq-modelwithAR1 minus @eq-constmean1) eliminates $\epsilon_{t}$ from the model:
$$
\begin{split}
Y_{t} - \phi Y_{t - 1} &= \left( \beta_{0} + \phi \epsilon_{t - 1} + a_{t} \right) - \left( \phi \beta_{0} + \phi \epsilon_{t - 1}  \right) \\
&= (1 - \phi) \beta_{0} + a_{t}.
\end{split}
$$

Therefore we can rewrite the constant mean model in @eq-constmean as
$$
Y_{t} = (1 - \phi) \beta_{0} + \phi Y_{t - 1} + a_{t}.
$${#eq-constmean2}

In general, for any multiple linear regression
$$
Y_{t} = \beta_{0} + \sum^{k}_{j =1} \beta_{j} X_{t,j} + \epsilon_{t}, ~~  \text{where} ~~ \epsilon_{t} \sim \mbox{AR(1)},
$${#eq-MlinModelAR1}
we can perform a similar procedure of eliminating $\epsilon_{t}$.

This elimination procedure leads to the alternate expression
$$
Y_{t} = (1 - \phi) \beta_{0} + \phi Y_{t -1} + \sum^{k}_{j = 1} \beta_{j} (X_{t,j} - \phi X_{t-1, j}) + a_{t},
$${#eq-MlinModelAR1alt}
where $a_{t}$ is white noise, i.e., homoskedastic with constant (zero) mean and uncorrelated. 
See @sec-gls describing the method of generalized least squares and an example of $k=1$.

Note that rewriting the model in this way pulls the autocorrelation parameter for the old residuals, $\phi$, into the regression part of the model.
Thus there are now $k + 2$ unknown regression parameters ($\beta_{0}, \beta_{1}, \dots, \beta_{k}$, and $\phi$). 
The introduction of an additional parameter into the regression part of the model can be regarded as the price to be paid for extracting new residuals $a_{t}$ that satisfy the OLS assumptions.

Note that the new residuals $a_{t}$ have smaller variance than the $\epsilon_{t}$. In fact,
$$
\begin{split}
\sigma^{2} & = \mbox{var} (\epsilon_{t} ) = \mbox{var} (\phi \epsilon_{t - 1} + a_{t}) \\
\\
& =  \phi^{2} \mbox{var}(\epsilon_{t - 1}) + \mbox{var} (a_{t} ) ~~~ \mbox{since} ~~ \mbox{cov}(a_{t}, \epsilon_{t - 1}) = 0\\
\\
 & = \phi^{2}\sigma^{2}  + \nu^{2},
\end{split}
$$
leading to the relation
$$
\nu^{2} = \sigma^{2} (1 - \phi^{2} ).
$${#eq-varresid}

However, comparing @eq-constmean2 with @eq-constmean2, and @eq-MlinModelAR1alt with @eq-MlinModelAR1, we see that the rewritten form of the model is *not linear* in terms of the parameters $\beta_{0}, \beta_{1}, \dots, \beta_{k}$ and $\phi$. 
For example, the intercept term $(1 - \phi) \beta_{0}$ involves a product of two of the parameters. 
This nonlinearity makes the OLS, implemented in the R functions `lm()` and `lsfit()`, a poor method for obtaining parameter estimates. 
Instead, we will use the method of *maximum likelihood* (ML) carried out through such R functions as `arima()`.

The function `arima()` allows us to input the model in its original form, as in @eq-MlinModelAR1. 
It then internally rewrites the model to put it in the form of @eq-MlinModelAR1alt. 
(So we do not have to rewrite the model!) It then makes the assumption that the $a_t$ are normal and constructs the multivariate normal likelihood function
$$
L (Y_{1} , \dots , Y_{n}; Q ),
$$
where $n$ is the sample size and $Q$ is the vector of all unknown parameters. 
In general, for an AR(1) model with $k$ original predictors, we have $Q = (\phi, \beta_{0}, \beta_{1}, \dots, \beta_{k}, \nu^{2})$. 
Recall that $\nu^{2} = \mathrm{var}(a_{t}) = \mathrm{cov}(a_{t} , a_{t})$.

The function `arima()` then uses the historical data $Y_{1}, \dots, Y_{n}$ to find the parameter estimates
$$
\hat{Q} = \left( \hat{\phi}, \hat{\beta}_{0} , \hat{\beta}_{1} , \dots, \hat{\beta}_{k} , \hat{\nu}^2 \right),
$$
which maximize the likelihood $L$. 
These estimates (and other things, such as the standard errors of the estimates) can be saved to an output object in R. 
We will use an example to illustrate the use and interpretation of the function `arima()`.

Moreover, we can extend the regression model in @eq-MlinModelAR1 with AR(1) errors to a model with a more general form of errors, ARMA, by assuming $\epsilon_t \sim \text{ARMA}(p,q)$ [see Chapter 6.6 in @Brockwell:Davis:2002 and <https://robjhyndman.com/hyndsight/arimax/>]:
$$
\begin{split}
Y_t &= \sum^{k}_{j =1} \beta_{j} X_{t,j} + \epsilon_{t},\\
\epsilon_{t} &= \phi_1 \epsilon_{t-1} + \dots + \phi_p \epsilon_{t-p} + \theta_1 a_{t-1} + \dots + \theta_q a_{t-q} + a_t.
\end{split}
$${#eq-MlinModelARMA}
This model in @eq-MlinModelARMA can be specified in R functions `arima()`, `fable::ARIMA()`, `forecast::Arima()`, and `forecast::auto.arima()`.

::: {.callout-note}
The R package `forecast` has been superseded by the new package `fable` that uses an alternate parameterization of constants, see `?fable::ARIMA`.
:::

::: {.callout-note}
Remember that the variables $Y_t$ and $X_{t,j}$ ($j = 1,\dots,k$) should be detrended prior to the analysis (to avoid spurious regression results, see the previous lecture on dealing with trends in regression). 
If differencing is chosen as the method of detrending, the orders of differences $d$ and $D$ can be specified directly within the mentioned ARIMA functions (so R will do the differencing for us).
:::


## ARIMAX models

ARIMAX ('X' stands for 'external regressor') models are closely related to @eq-MlinModelARMA, but there is an important difference. 
For simplicity of notation, we can present an ARMAX($p,q$) model that is a regular ARMA($p,q$) model for $Y_t$ plus the external regressors:
$$
\begin{split}
Y_t &= \phi_1 Y_{t-1} + \dots + \phi_p Y_{t-p} + \theta_1 a_{t-1} + \dots + \theta_q a_{t-q} + a_t\\
&+\sum^{k}_{j =1} \beta_{j} X_{t,j},
\end{split}
$${#eq-ARMAX}
where $a_t$ is still a zero-mean white noise process. 
Interestingly, at this time there is no convenient way to estimate this model in R. 
One could *manually* write lagged values of $Y_t$ as external regressors (i.e., create new variables in R for $Y_{t-1},\dots, Y_{t-p}$), use these variables in the R functions mentioned above, but force $p=0$ [e.g., @Soliman:etal:2019:influenza used the functions from the R package `forecast`].

The difference between @eq-MlinModelARMA and @eq-ARMAX is the presence of lagged values of the response variable, $Y_t$, in @eq-ARMAX. 
As Hyndman points out, regression coefficients $\beta_j$ in @eq-ARMAX lose their interpretability compared with usual regression and do not show the effect on $Y_t$ when $X_t$ increased by one. 
Instead, $\beta$'s in ARMAX @eq-ARMAX are interpreted *conditional* on the value of previous values of the response variable (<https://robjhyndman.com/hyndsight/arimax/>).
Therefore, model formulation as in @eq-MlinModelARMA may be preferred.

::: {.callout-note}
Other applicable models include models with mixed effects such as for *repeated measures ANOVA* (e.g., implemented in R using `nlme::gls()` and `nlme::lme()`, see different correlation structures; without a grouping factor, the results of `nlme::lme(..., correlation = corARMA(...))` should be similar to estimating model in @eq-MlinModelARMA). 
Other regression functions often borrow the functionality (and syntax) of the package `nlme` for estimating random effects, so autocorrelated residuals can be incorporated into a *generalized additive model*, GAM, `mgcv::gamm()`; *generalized additive model for location scale and shape*, GAMLSS,  `gamlss::gamlss()`, which also can be used just as GAM, if scale and shape parameters are not modeled. 
A slightly different solution is possible using a *generalized autoregressive moving average model* (GARMA) demonstrated in @sec-GARMA, see `gamlss.util::garmaFit()`.
:::

::: {.callout-note icon=false}

## Example: Golden tilefish ARIMAX

Here we have time series of 1918--2017 annual landings of golden tilefish (tonne) in the U.S. North Atlantic region and the Atlantic Multi-decadal Oscillation (AMO) index characterizing climatic conditions (@fig-tilefishTS). 
These time series were described in @Nesslage:etal:2021:environmental and @lyubchich3732840.
The goal is to develop a regression model to explore the relationship between the landings and AMO.

```{r}
D <- read.csv("data/tilefish.csv")
summary(D)
```

```{r}
#| label: fig-tilefishTS
#| fig-cap: "Time series plots of the golden tilefish landings and AMO."

p1 <- ggplot(D, aes(x = Year, y = Landings)) +
    geom_line() +
    xlab("Year") +
    ylab("Landings (tonne)") +
    theme_light()
p2 <- ggplot(D, aes(x = Year, y = AMO)) +
    geom_line() +
    xlab("Year") +
    ylab("AMO") +
    theme_light()
p1 + p2 +
    plot_annotation(tag_levels = 'A')
```

Interpolation of missing data (a.k.a. imputation) is a separate and very expansive topic. For a univariate time series, linear interpolation can be implemented using `forecast::na.interp()`. 
Also, the landings time series required a power transformation, so square root transformation was applied (@fig-tilefishScatter).

```{r}
#| label: fig-tilefishScatter
#| fig-cap: "Scatterplot matrix of landings and AMO."

D <- D %>% 
    mutate(Landings_noNA = as.numeric(forecast::na.interp(D$Landings))) %>% 
    mutate(Landings_noNA_sqrt = sqrt(Landings_noNA))

D %>% 
    select(AMO, Landings_noNA_sqrt) %>% 
    GGally::ggpairs() +
    theme_light()
```

The lag `k` implemented in `ccf(x, y)` is used to estimate the correlation between `x[t + k]` and `y[t]` (@fig-tilefishCCF).

```{r}
#| label: fig-tilefishCCF
#| fig-cap: "Estimated cross-correlation function (CCF) of time series of the landings and AMO."

par(mar = c(4, 5, 1, 1) + 0.1, mgp = c(3, 1, 0), mfrow = c(1, 1))
with(D,
     ccf(Landings_noNA_sqrt, AMO, las = 1)
)
```

Implement a model
$$
\sqrt{Landings_t} = \beta_0 + \beta_{1} AMO_{t-7} + \epsilon_{t},
$$
where $\epsilon_{t} \sim$ ARMA($p,q$), and the orders $p$ and $q$ are selected automatically based on Akaike information criterion.

```{r}
library(fable)
m1 <- D %>% 
    select(Year, Landings_noNA_sqrt, AMO) %>% 
    as_tsibble(index = Year) %>% 
    model(ARIMA(Landings_noNA_sqrt ~ dplyr::lag(AMO, 7)))

report(m1)
```

We forgot to check the stationarity of the time series! 
It seems that the landings can be considered as a unit-root process, so differencing is needed, hence a modified model is

$$
\Delta \sqrt{Landings_t} = \beta_0 + \beta_{1} AMO_{t-7} + \epsilon_{t}.
$$

```{r}
m2 <- forecast::auto.arima(diff(D$Landings_noNA_sqrt),
                 xreg = dplyr::lag(D$AMO, 7)[-1],
                 allowmean = TRUE)
m2
```

Note it is different from the implementation with `d = 1`, which differences all the series.

$$
\Delta \sqrt{Landings_t} = \beta_0 + \beta_{1} \Delta AMO_{t-7} + \epsilon_{t}
$$
```{r}
forecast::auto.arima(D$Landings_noNA_sqrt,
                     xreg = dplyr::lag(D$AMO, 7), 
                     d = 1)
```

Similarly, using the newer package `fable`.

```{r}
m3 <- D %>% 
    select(Year, Landings_noNA_sqrt, AMO) %>% 
    as_tsibble(index = Year) %>% 
    model(ARIMA(Landings_noNA_sqrt ~ pdq(d = 1) + dplyr::lag(AMO, 7)))

report(m3)
```
:::


## GARMA models {#sec-GARMA}

Let $Y_t$ be the observed time series and $\boldsymbol{X_t}$ exogenous regressors. 
Then, we can model the conditional distribution of $Y_t$, given $Y_1,\dots,Y_{t-1}$, $\boldsymbol{X_1,\dots,X_{t}}$ as
$$
g(\mu_t)=\boldsymbol{X}'_t\beta+\sum_{j=1}^p\phi_j\{g(Y_{t-j})- \boldsymbol{X}'_{t-j}\beta\}
+\sum_{j=1}^q\theta_j \{g(Y_{t-j})-g(\mu_{t-j})\},
$${#eq-GARMA}
where $g(\cdot)$ is an appropriate link function; $\mu_t$ is a conditional mean of the dependent variable; $\boldsymbol{\beta}$ is a vector of regression coefficients; $\phi_j$, $j=1,\dots, p$, are the autoregressive coefficients; $\theta_j$, $j=1,\dots,q$, are the moving average coefficients, while $p$ and $q$ are the autoregressive and moving average orders, respectively.

In certain cases, the function $g(\cdot)$ requires some transformation of the original series $Y_{t-j}$ to avoid the non-existence of $g(Y_{t-j})$ [@Benjamin:etal:2003].

The generalized autoregressive moving average model (@eq-GARMA), GARMA($p,q$), represents a flexible observation-driven modification of the classical Box--Jenkins methodology and GLMs for integer-valued time series. 
GARMA further advances the classical Gaussian ARMA model to a case where the distribution of the dependent variable is not only non-Gaussian but can be discrete. 
The dependent variable is assumed to belong to a conditional exponential family distribution given the past information of the process and thus the GARMA can be used to model a variety of discrete distributions [@Benjamin:etal:2003]. 
The GARMA model also extends the work of @Zeger:Qaqish:1988, who proposed an autoregressive exponential family model, and @Li:1994, who introduced its moving average counterpart.

:::{.callout-note icon=false}

## Example: Insurance claims GARMA model

Consider the weekly number of house insurance claims related to water and weather damage in one Canadian city. 
The number is standardized by the daily number of insured properties in that city. 
Explore the relationship between the number of claims and weekly total precipitation (mm).

```{r}
Insurance <- read.csv("data/insurance_weekly.csv") %>%
    select(Claims, Precipitation)
summary(Insurance)
```

```{r}
#| label: fig-insScatter
#| fig-cap: "Scatterplot matrix of weekly weather-related home insurance outcomes and precipitation."

Insurance %>%
    GGally::ggpairs() +
    theme_light()
```

Based on the distribution plots in @fig-insScatter, the data are highly right-skewed (have heavy right tails). 
The number of claims is also a discrete variable. 
Therefore, we deal with non-normal distributions and need to use generalized-type models, like the generalized linear or additive models (GLMs or GAMs).
Since there are many zeros in the counts of claims, we can start with the zero-adjusted Poisson distribution to model the number of claims [@Stasinopoulos:Rigby:2007;@Gupta:etal:1996].

In our case, there is just a slight chance that past-week precipitation affects the current-week insurance claims. 
Hence, we will still keep the current precipitation and additionally explore the lagged effects, using the cross-correlation function (@fig-insCCF).

```{r}
#| label: fig-insCCF
#| fig-cap: "Estimated cross-correlation function (CCF) of time series of the number of home insurance claims and precipitation."

par(mar = c(4, 5, 1, 1) + 0.1, mgp = c(3, 1, 0), mfrow = c(1, 1))
logconstant <- 1
with(Insurance,
     ccf(log(Claims + logconstant), Precipitation, las = 1, lag.max = 3)
)
```

Based on the estimated CCFs (@fig-insCCF), past-week precipitation is significantly correlated with the current-week number of claims, so we can add the lagged predictor into our models.

```{r}
Insurance <- Insurance %>%
    mutate(Precipitation_lag1 = dplyr::lag(Precipitation, 1),
           Week = 1:nrow(Insurance),
           Claims_ln = log(Claims + logconstant))
```

Based on @fig-insACF, there might be an increasing trend in the number of claims that we might be able to approximate with a linear function.

```{r}
#| label: fig-insACF
#| fig-cap: "Time series plot and sample ACF of the weekly number of claims in a city, after the ln transformation."

p1 <- ggplot2::autoplot(as.ts(log(Insurance$Claims + logconstant))) +
    xlab("Week") +
    ylab("ln(Number of claims)") +
    theme_light()
p2 <- forecast::ggAcf(as.ts(log(Insurance$Claims + logconstant)),
                      lag.max = 110) +
    ggtitle("") +
    xlab("Lag (weeks)") +
    theme_light()
p1 + p2 +
    plot_annotation(tag_levels = 'A')
```

Plot the data once again after the transformations (@fig-insScatter2).

```{r}
#| label: fig-insScatter2
#| fig-cap: "Scatterplot matrix of the weekly number of weather-related home insurance claims and precipitation."

Insurance %>%
    dplyr::select(-Claims) %>%
    GGally::ggpairs() +
    theme_light()
```

Fit a GARMA model relating the weekly number of insurance claims to the total precipitation during that and previous weeks.

```{r}
#| code-fold: false

# The model function doesn't accept NAs, so remove them
Insurance_noNA <- na.omit(Insurance)

library(gamlss.util)
m00zip <- garmaFit(Claims ~ Precipitation + Week + Precipitation_lag1
                   ,family = ZIP
                   ,data = Insurance_noNA)
```

Obtain ACF and PACF plots of the model residuals to select ARMA order (@fig-insresACF00).

```{r}
#| label: fig-insresACF00
#| fig-cap: "ACF and PACF plots of residuals of the base model GARMA(0,0) based on ZIP distribution."

p1 <- forecast::ggAcf(m00zip$residuals) +
    ggtitle("") +
    xlab("Lag (weeks)") +
    theme_light()
p2 <- forecast::ggAcf(m00zip$residuals, type = "partial") +
    ggtitle("") +
    xlab("Lag (weeks)") +
    theme_light()
p1 + p2 +
    plot_annotation(tag_levels = 'A')
```

Based on the observed ACF and PACF patterns in @fig-insresACF00, an appropriate model for the temporal dependence could be ARMA(3,0). 
Refit the GARMA model specifying these orders. 
Then verify that the temporal dependence in residuals was removed (@fig-m03zipACF), and assess other assumptions (@fig-m03zip), including homogeneity and normality of the quantile residuals.

```{r}
#| code-fold: false

set.seed(12345)
m30zip <- garmaFit(Claims ~ Precipitation + Week + Precipitation_lag1
                ,order = c(3, 0)
                ,family = ZIP
                ,data = Insurance_noNA)
summary(m30zip)
```

```{r}
#| label: fig-m03zipACF
#| fig-cap: "ACF and PACF plots of the GARMA(0,3) model residuals based on ZIP distribution."

p1 <- forecast::ggAcf(m30zip$residuals) +
    ggtitle("") +
    xlab("Lag (weeks)") +
    theme_light()
p2 <- forecast::ggAcf(m30zip$residuals, type = "partial") +
    ggtitle("") +
    xlab("Lag (weeks)") +
    theme_light()
p1 + p2 +
    plot_annotation(tag_levels = 'A')
```


```{r}
#| label: fig-m03zip
#| fig-cap: "Default diagnostics of the GARMA(0,3) model residuals based on ZIP distribution."
#| fig-height: 7

plot(m30zip)
```

See @Dunn:Smyth:1996 for the details on randomized quantile residuals. 
Overall, their poor correspondence with the standard normal distribution shows a lack of fit of the model. 
See `?gamlss.family` for other distribution families, continuous and discrete; somewhat out-of-date tables with many of these distributions listed are available from @Stasinopoulos:Rigby:2007.

Here we try one other distribution appropriate for modeling overdispersed count data -- negative binomial distribution. 
See the model summary below and the verification of residuals in @fig-m03nbiACF and @fig-m03nbi.

```{r}
#| code-fold: false

set.seed(12345)
m03nbi <- garmaFit(Claims ~ Precipitation + Week + Precipitation_lag1
                ,order = c(0, 3)
                ,family = NBI
                ,data = Insurance_noNA)
summary(m03nbi)
```

```{r}
#| label: fig-m03nbiACF
#| fig-cap: "ACF and PACF plots of the GARMA(0,3) model residuals based on NBI distribution."

p1 <- forecast::ggAcf(m03nbi$residuals) +
    ggtitle("") +
    xlab("Lag (weeks)") +
    theme_light()
p2 <- forecast::ggAcf(m03nbi$residuals, type = "partial") +
    ggtitle("") +
    xlab("Lag (weeks)") +
    theme_light()
p1 + p2 +
    plot_annotation(tag_levels = 'A')
```


```{r}
#| label: fig-m03nbi
#| fig-cap: "Default diagnostics of the GARMA(0,3) model residuals based on NBI distribution."
#| fig-height: 7

plot(m03nbi)
```

The residual diagnostics look better for the latter model. 
One could also consider modeling the strictly periodical component such as using the Fourier series, however, @fig-insACF did not show a prominent seasonality.
:::


## Granger causality {#sec-Granger}

The Granger causality [@Granger:1969;@Kirchgassner:Wolters:2007] concept is based on predictability, which is why we should consider it in the time series course.
Pearl causality is based on the analysis of interventions [@Rebane:Pearl:1987;@Pearl:2009].

Let $I_t$ be the total information set available at the time $t$. 
This information set includes the two time series $X$ and $Y$. 
Let $\bar{X}_t$ be the set of all current and past values of $X$, i.e., $\bar{X}_t = \{X_{t}, X_{t-1}, \dots, X_{t-k}, \dots \}$ and analogously of $Y$. Let $\sigma^2(\cdot)$ be the variance of the corresponding forecast error.

::: {.callout-note}
Difference of two sets, $A$ and $B$, is denoted by $A \setminus B$; but sometimes the minus sign is used, $A - B$.
:::

**Granger causality**

$X$ is (simply) Granger causal to $Y$ if future values of $Y$ can be predicted better, i.e., with a smaller forecast error variance, if current and past values of $X$ are used:
$$
\sigma^2(Y_{t+1}|I_t) < \sigma^2(Y_{t+1}|I_t \setminus \bar{X}_t).
$${#eq-Granger}

**Instantaneous Granger causality**

$X$ is instantaneously Granger causal to $Y$ if the future value of $Y$, $Y_{t+1}$, can be predicted better, i.e., with a smaller forecast error variance, if the future value of $X$, $X_{t+1}$, is used in addition to the current and past values of $X$:
$$
\sigma^2(Y_{t+1}|\{I_t, X_{t+1}\}) < \sigma^2(Y_{t+1}|I_t ).
$${#eq-GrangerInst}

**Feedback**

There is feedback between $X$ and $Y$ if $X$ is causal to $Y$ and $Y$ is causal to $X$. 
Feedback is only defined for the case of simple causal relations.

The test for @eq-Granger and @eq-GrangerInst is, essentially, an $F$-test comparing two nested models: with additional predictors $X$ and without. 
In other words, consider the model:
$$
Y_t = \beta_0 + \sum_{k=1}^{k_1}\beta_k Y_{t-k} + \sum_{k=k_0}^{k_2}\alpha_k X_{t-k} + U_t
$${#eq-GrangerModel}
with $k_0 = 1$. 
An $F$-test is applied to test the null hypothesis, H$_0$: $\alpha_1 = \alpha_2 = \dots = \alpha_{k_2} = 0$. 
By switching $X$ and $Y$ in @eq-GrangerModel, it can be tested whether a simple causal relation from $Y$ to $X$ exists. 
There is a feedback relation if the null hypothesis is rejected in both directions ($X\rightarrow Y$ and $Y\rightarrow X$). 
To test whether there is an instantaneous causality, we finally set $k_0 = 0$ and perform a $t$ or $F$-test for the null hypothesis H$_0$: $\alpha_0 = 0$.

The problem with this test is that the results are strongly dependent on the number of lags of the explanatory variable, $k_2$. 
There is a trade-off: the more lagged values we include, the better the influence of this variable can be captured. 
This argues for a high maximal lag. On the other hand, the power of this test is lower the more lagged values are included [Chapter 3 of @Kirchgassner:Wolters:2007]. 
Two general procedures can be used to select the lags: inspecting the sensitivity of results to different $k_2$ (sensitivity analysis) or one of the different information criteria guiding model selection.

::: {.callout-note icon=false}

## Example: Insurance claims and precipitation Granger causality test

For example, use the insurance data to test the relationships between the log-transformed home insurance number of claims and precipitation. 
A quick test has been performed to check that the time series are stationary and do not have a strong seasonality, so the chance of detecting spurious relationships is minimized. 
We can use the series in @eq-GrangerModel.

```{r}
#| code-fold: false

lmtest::grangertest(Claims_ln ~ Precipitation 
                    ,order = 1
                    ,data = Insurance_noNA)
```

The low $p$-value shows that precipitation is a Granger cause of the number of home insurance claims. 
Now test the reverse relationship.

```{r}
#| code-fold: false

lmtest::grangertest(Precipitation ~ Claims_ln  
                    ,order = 1
                    ,data = Insurance_noNA)
```

Reverse testing does not confirm a statistically significant Granger causality. 
Hence, we do not have enough evidence to claim that insurance claims are a Granger cause of precipitation (we could expect to come to this conclusion based on our knowledge of how things work), and hence there is no evidence of feedback between losses and precipitation.

The function `lmtest::grangertest()` does not allow us to set `order = 0` (to test instantaneous Granger causality), but we can do it manually. 
First, to show that 'manual' results match the `lmtest::grangertest()` output, repeat the test above using two nested models with lag 1.

```{r}
#| code-fold: false

M1 <- lm(Claims_ln ~ dplyr::lag(Claims_ln, 1) + dplyr::lag(Precipitation, 1)
         ,data = Insurance_noNA)
M2 <- lm(Claims_ln ~ dplyr::lag(Claims_ln, 1)
         ,data = Insurance_noNA)
anova(M1, M2)
```

The results match.

Second, test the instantaneous Granger causality in the same manner.

```{r}
#| code-fold: false

M3 <- lm(Claims_ln ~ dplyr::lag(Claims_ln, 1) + dplyr::lag(Precipitation, 1) + 
             Precipitation, data = Insurance_noNA)
M4 <- lm(Claims_ln ~ dplyr::lag(Claims_ln, 1) + dplyr::lag(Precipitation, 1)
         ,data = Insurance_noNA)
anova(M3, M4)
```

The results show the presence of such causality.

As an extension to the demonstrated techniques for testing Granger causality, we may consider more complex generalized and additive models, to relax the assumptions of normality and linearity in the modeling process.
:::

::: {.callout-note}
The `lmtest::grangertest()` options set one value to both $k_1$ and $k_2$ in @eq-GrangerModel. 
In our example, it was $k_1 = k_2 = 1$. 
The 'manual' test using the function `anova()` can be used for models with $k_1 = k_2$ or $k_1 \neq k_2$.
:::


## About predictions

Let $\hat{Y}_T(h)$ be a forecast $h$ steps ahead made at the time $T$. 
If $\hat{Y}_T(h)$ only uses information up to time $T$, the resulting forecasts are called out-of-sample forecasts. 
Economists call them *ex-ante* forecasts. 
We have discussed several ways to select the optimal method or model for forecasting, e.g., using PMAE, PMSE, or coverage -- all calculated on a testing set. 
@Chatfield:2000 mentions several ways to unfairly 'improve' forecasts:

1. Fitting the model to all the data including the test set.
1. Fitting several models to the training set and choosing the model which gives the best 'forecasts' of the test set. 
The selected model is then used (again) to produce forecasts of the test set, even though the latter has already been used in the modeling process.
1. Using the known test-set values of 'future' observations on the explanatory variables in multivariate forecasting. 
This will improve forecasts of the dependent variable in the test set, but these future values will not of course be known at the time the forecast is supposedly made (though in practice the 'forecast' is made at a later date). 
Economists call such forecasts *ex-post* forecasts to distinguish them from *ex-ante* forecasts. 
The latter, being genuinely out-of-sample, use forecasts of future values of explanatory variables, where necessary, to compute forecasts of the response variable. 
*Ex-post* forecasts can be useful for assessing the effects of explanatory variables, provided the analyst does not pretend that they are genuine out-of-sample forecasts.

So what to do if we put lots of effort to build a regression model using time series and need to forecast the response, $Y_t$, which is modeled using different independent variables $X_{t,k}$ ($k=1,\dots,K$)? 
Two options are possible.

**Leading indicators**

If $X_{t,k}$'s are leading indicators with lags starting at $l$, we, generally, would not need their future values to obtain the forecasts $\hat{Y}_T(h)$, where $h\leqslant l$. 
For example, the model for losses tested in @sec-Granger shows that precipitation with lag 1 is a good predictor for current losses, i.e., precipitation is a leading indicator. 
The 1-week ahead forecast of $Y_{t+1}$ can be obtained using the current precipitation $X_t$ (all data are available). 
If $h>l$, we will be forced to forecast the independent variables, $X_{t,k}$'s -- see the next option.

**Forecast of predictors**

If we opt for forecasting $X_{t,k}$'s, the errors (uncertainty) of such forecasts will be larger, because future $X_{t,k}$'s themselves will be the estimates.
Nevertheless, it might be the only choice when leading indicators are not available. 
Building a full and comprehensive model with all diagnostics for each regressor is usually unfeasible and even problematic if we plan to consider multivariate models for regressors (the complexity of models will quickly escalate). 
As an alternative, it is common to use automatic or semi-automatic univariate procedures that can help to forecast each of the $X_{t,k}$'s. 
For example, consider exponential smoothing, Holt--Winters smoothing, and auto-selected SARIMA/ARIMA/ARMA/AR/MA models -- all those can be automated for a large number of forecasts to make.


## Conclusion

Multivariate models are still much more difficult to fit than univariate ones. 
Multiple regression remains a treacherous procedure when applied to time series data. 
Many observed time series exhibit nonlinear characteristics, but nonlinear models often fail to give better out-of-sample forecasts than linear models, perhaps because the latter are more robust to departures from model assumptions. 
It is always a good idea to end with the so-called eyeball test. 
Plot the forecasts on a time plot of the data and check that they look intuitively reasonable [@Chatfield:2000].
