---
output: html_document
editor_options:
    chunk_output_type: console
---

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
library(patchwork)
options(digits = 3)
```

# Generalized least squares {#sec-gls}

Here we use time series data (ordered by $t$), thus, @eq-mod1 will be written with the time indices $t$ as
$$
y_t=\beta_0+\beta_1x_t+\varepsilon_t,
$${#eq-mod1t}
where the regression errors at times $t$ and $t-1$ are
$$
\begin{split}
\varepsilon_t&=y_t-\beta_0-\beta_1x_t,\\
\varepsilon_{t-1}&=y_{t-1}-\beta_0-\beta_1x_{t-1}.
\end{split}
$${#eq-mod1error}

An AR(1) model for the errors will yield
$$
\begin{split}
y_t-\beta_0-\beta_1x_t & = \rho\varepsilon_{t-1} + w_t, \\
y_t-\beta_0-\beta_1x_t & = \rho(y_{t-1}-\beta_0-\beta_1x_{t-1})+w_t,
\end{split}
$${#eq-modAR1}
where $w_t$ are uncorrelated errors.

Rewrite it as
$$
\begin{split}
y_t-\rho y_{t-1}&=\beta_0(1-\rho)+\beta_1(x_t-\rho x_{t-1})+w_t,\\
y_t^* &= \beta_0^* + \beta_1 x_t^*+w_t,
\end{split}
$${#eq-mod1w}
where $y_t^* = y_t-\rho y_{t-1}$; $\beta_0^* = \beta_0(1-\rho)$; $x_t^* = x_t-\rho x_{t-1}$. 
Notice the errors $w_t$ in the final @eq-mod1w for the transformed variables $y_t^*$ and $x_t^*$ are uncorrelated.

To get from @eq-mod1t to @eq-mod1w, we can use an iterative procedure by @Cochrane:Orcutt:1949 as in the example below.

::: {.callout-note icon=false}

## Example: Dishwasher shipments model accounting for autocorrelation

1. Estimate the model in @eq-mod1t using OLS.
```{r}
D <- read.delim("data/dish.txt") %>% 
    rename(Year = YEAR)
modDish_ols <- lm(DISH ~ RES, data = D)
```

2. Calculate residuals $\hat{\varepsilon}_t$ and estimate $\rho$ as
$$
\hat{\rho}=\frac{\sum_{t=2}^n\hat{\varepsilon}_t\hat{\varepsilon}_{t-1}}{\sum_{t=1}^n\hat{\varepsilon}^2_t}.
$$
```{r}
e <- modDish_ols$residuals
rho <- sum(e[-1] * e[-length(e)]) / sum(e^2)
rho
```

3. Calculate transformed variables $x^*_t$ and $y^*_t$ and fit model in @eq-mod1w.
```{r}
y.star <- D$DISH[-1] - rho * D$DISH[-length(D$DISH)]
x.star <- D$RES[-1] - rho * D$RES[-length(D$RES)]
modDish_ar1 <- lm(y.star ~ x.star)
summary(modDish_ar1)
```

4. Examine the residuals of the newly fitted equation (@fig-dishAR1) and repeat the procedure, if needed.

```{r}
#| label: fig-dishAR1
#| fig-cap: "Residual plots of the original OLS model and the model transformed to account for autocorrelation in residuals."

p1 <- ggplot(D, aes(x = Year, y = modDish_ols$residuals)) + 
    geom_line() + 
    geom_hline(yintercept = 0, lty = 2, col = 4) + 
    ggtitle("OLS model modDish_ols") +
    ylab("Residuals") +
    theme_light()
p2 <- ggplot(D[-1,], aes(x = Year, y = modDish_ar1$residuals)) + 
    geom_line() + 
    geom_hline(yintercept = 0, lty = 2, col = 4) + 
    ggtitle("Transformed model modDish_ar1") +
    ylab("Residuals") +
    theme_light()
p1 + p2 +
    plot_annotation(tag_levels = 'A')
```

Based on the runs test, there is not enough evidence of autocorrelation in the new residuals:

```{r}
#| code-fold: false

lawstat::runs.test(rstandard(modDish_ar1))
```
:::

What we have just applied is the method of *generalized least squares* (GLS):
$$
\hat{\boldsymbol{\beta}} = \left( \boldsymbol{X}^{\top}\boldsymbol{\Sigma}^{-1}\boldsymbol{X}\right)^{-1} \boldsymbol{X}^{\top}\boldsymbol{\Sigma}^{-1}\boldsymbol{Y},
$${#eq-gls}
where $\boldsymbol{\Sigma}$ is the covariance matrix. 
The method of weighted least squares (WLS; @sec-wls) is just a special case of the GLS. 
In the WLS approach, all the off-diagonal entries of $\boldsymbol{\Sigma}$ are 0.

We can use the function `nlme::gls()` and specify the `correlation` structure to avoid iterating the steps from the previous example manually:

```{r}
#| code-fold: false

modDish_ar1_v2 <- nlme::gls(DISH ~ RES
                            ,correlation = nlme::corAR1(form = ~Year)
                            ,data = D)
summary(modDish_ar1_v2)
```

::: {.callout-note}
In the function `nlme::gls()` we can also specify `weights` to accommodate heteroskedastic errors, but the syntax differs from the `weights` specification in the function `stats::lm()` (@sec-wls). 
See `?nlme::varFixed`.
:::
