---
output: html_document
editor_options:
    chunk_output_type: console
---

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
library(patchwork)
options(digits = 3)
```

# Synchrony of parametric trends {#sec-syncparam}

The problem of detecting joint trend dynamics in time series is essential in a variety of applications, ranging from the analysis of macroeconomic indicators [@Vogelsang:Franses:2005;@Eun:Lee:2010] to assessing patterns in ice phenology measurements from multiple locations [@Latifovic:Pouliot:2007;@Duguay:etal:2013] to evaluating yields of financial instruments at various maturity levels [@Park:etal:2009] and cell phone download activity at different area codes [@Degras:etal:2012].

The extensive research on comparing trend patterns follows two main directions: 

1. Testing for joint mean functions.
1. Analysis of joint stochastic trends, which is closely linked to the cointegration notion by @Engle:Granger:1987 (@sec-cointegration).

Here we explore the first direction, that is, assess whether several observed time series follow the same hypothesized parametric trend. 

There exist many tests for comparing mean functions, but most of the developed methods assume independent errors. 
Substantially less is known about testing for joint deterministic trends in a time series framework.

One of the methods developed for time series is by @Degras:etal:2012 and @Zhang:2013 who extended the integrated square error (ISE) based approach of @VF:GM:2004 to a case of multiple time series with weakly dependent (non)stationary errors. 
For a comprehensive literature review of available methodology for comparing mean functions embedded into independent errors in a time series framework, see @Degras:etal:2012 and @Park:etal:2014. 
Most of these methods, however, either focus on aligning only two curves or require us to select multiple hyperparameters, such as the bandwidth, level of smoothness, and window size for a long-run variance function. 
As mentioned by @Park:etal:2014, the choice of such *multiple nuisance parameters* is challenging for a comparison of curves (even under the independent and identically distributed setup) and often leads to inadequate performance, especially in samples of moderate size.

As an alternative, consider an extension of the WAVK test (@sec-WAVK) to a case of multiple time series [@Lyubchich:Gel:2016:synchronism]. 
Let us observe $N$ time series
$$
Y_{it} = \mu_i(t/T) + \epsilon_{it},
$$
where $i = 1, \dots, N$ ($N$ is the number of time series), $t=1, \dots, T$ ($T$ is the length of the time series), $\mu_i(u)$ ($0<u\leqslant 1$) are unknown smooth trend functions, and the noise $\epsilon_{it}$ can be represented as a finite-order AR($p$) process or infinite-order AR($\infty$) process with i.i.d. innovations ${e}_{it}$.

We are interested in testing whether these $N$ observed time series have the same trend of some pre-specified smooth parametric form $f(\theta, u)$:

$H_0$: $\mu_i(u)= c_i + f(\theta, u)$  
$H_1$: there exists $i$, such that $\mu_i(u)\neq c_i + f(\theta, u)$,

where the reference curve $f(\theta, u)$ with a vector of parameters $\theta$ belongs to a known family of smooth parametric functions, and $1\leqslant i \leqslant N$. 
For identifiability, assume that $\sum_{i=1}^N c_i=0$. 
Notice that the hypotheses include (but are not limited to) the special cases of  
$f(\theta,u)\equiv 0$ (testing for no trend);  
$f(\theta,u)=\theta_0+\theta_1 u$ (testing for a common linear trend);  
$f(\theta,u)=\theta_0+\theta_1 u+\theta_2u^2$ (testing for a common quadratic trend).

This hypothesis testing approach allows us to answer the following questions:

- Do trends in temperature (or wind speeds, or precipitation) reproduced by a climate model correspond to the historical observations? 
I.e., is the model generally correct? 
- Do different instruments (sensors) capture changes similarly, or deviate, for example, due to aging of some of the instruments?
- Do trends estimated at different locations (Canada and USA, lower and mid-troposphere, etc.) follow some hypothesized global trend?

Test the null hypothesis by following these steps [@Lyubchich:Gel:2016:synchronism]:

1. Estimate the joint hypothetical trend $f({\theta}, u)$ using the aggregated sample $\left\{\overline{Y}_{\cdot t}\right\}_{t=1}^T$ (i.e., a time series obtained by averaging across all $N$ time series).
1. For each time series, subtract the estimated trend, then apply the autoregressive filter to obtain residuals $\hat{e}_{it}$, which under the $H_0$ behave asymptotically like the independent and identically distributed ${e}_{it}$:
$$
\begin{split}
\hat{e}_{it}&= \hat{\epsilon}_{it}-\sum_{j=1}^{p_i(T)}{\hat{\phi}_{ij}\hat{\epsilon}_{i,t-j}} \\
&=
\left\{ Y_{it} - f(\hat{\theta},u_{t}) \right\} -
\left\{ \sum_{j=1}^{p_i(T)}{\hat{\phi}_{ij}{Y}_{i,t-j}} - \sum_{j=1}^{p_i(T)}{\hat{\phi}_{ij}f(\hat{\theta},u_{t-j})} \right\}.
\end{split}
$$
1. Construct a sequence of $N$ statistics ${\rm WAVK}_{1}(k_{1}), \dots, {\rm WAVK}_{N}(k_{N})$.
Then, the synchrony test statistic is
$$
S_T = \sum_{i=1}^N k_{i}^{-1/2}{\rm WAVK}_{i}(k_{i}),
$$
where $k_{i}$ is the local window size for the WAVK statistic.
1. Estimate the variance of $\hat{e}_{it}$, e.g., using the robust difference-based estimator by @Rice:1984:
$$
s_i^2= \frac{\sum_{t=2}^T(\hat{e}_{it}-\hat{e}_{i,t-1})^2}{2(T-1)}.
$$
1. Simulate $BN$ times $T$-dimensional vectors $e^*_{iT}$ from the multivariate normal distribution $MVN\left(0, s_i^2\boldsymbol{I}\right)$, where $B$ is the number of bootstrap replications, $\boldsymbol{I}$ is a $T\times T$ identity matrix.
1. Compute $B$ bootstrapped statistics on $e^*_{iT}$:
$$S^*_T=\sum_{i=1}^N k^{-1/2}_{i} {\rm WAVK}^*_{i}(k_{i}).$$
1. The bootstrap $p$-value for testing the $H_0$ is the proportion of $|S^*_T|$ that exceed $|S_T|$.

See the application of both the WAVK and synchrony tests in @Lyubchich:2016:trends.

If the null hypothesis is rejected, the method does not tell, however, what was the reason, and which particular time series caused the rejection of the $H_0$.
One can remove the time series (or several time series at once) with the largest WAVK statistic(s) and apply the test again, although repeated testing increases the probability of Type I error. 
For an application of this method in trend clustering, see [this vignette](https://cran.r-project.org/web/packages/funtimes/vignettes/tsclusters.html).

::: {.callout-note icon=false}

## Example: CMIP5 vs. observations

Replicate the test for synchrony of trends found in two time series [@Lyubchich:2016:trends]:

- a multi-model average of temperatures from the 5th phase of the Coupled Model Intercomparison Project (CMIP5) and
- observed global temperature anomalies relative to the base period of 1981--2010.

```{r}
D <- read.csv("data/CMIP5.csv") %>% 
    filter(1948 <= Year & Year <= 2013) %>%
    mutate(Temp_CMIP5 = Temp_CMIP5 - 273.15)
```

See @fig-CMIP5 showing the time series plots after converting the CMIP data to degrees Celsius.

```{r}
#| label: fig-CMIP5
#| fig-cap: "Global annual mean temperature (°C) in 1948--2013: CMIP5 multi-model average and observed anomalies relative to the base period of 1981--2010."

p1 <- D %>% ggplot(aes(x = Year, y = Temp_CMIP5)) +
    geom_line() +
    theme_light()
p2 <- D %>% ggplot(aes(x = Year, y = Temp_obs)) +
    geom_line() +
    theme_light()
p1 + p2 +
    plot_annotation(tag_levels = 'A')
```

Test the synchrony of parametric *linear* trends in these time series:

```{r}
#| code-fold: false

set.seed(123)
funtimes::sync_test(D[, c("Temp_CMIP5", "Temp_obs")] ~ t)
```

The $p$-value below the usual significance level $\alpha = 0.05$ allows us to reject the null hypothesis, however, as @Lyubchich:2016:trends pointed out, the decision would differ if more confidence is required (e.g., $\alpha = 0.01$). 
Note that the $p$-value of 0.012 reported by @Lyubchich:2016:trends differs from the one reported above due to the function settings and randomness due to the bootstrapping.
We should save the random number generator state with `set.seed()` for replicability (so the test results are exactly the same every time the test is applied), and use a larger number of bootstrap replications `B` for consistency (so the test leads to the same conclusions when the function `set.seed()` is not used).

Now test the synchrony of parametric *quadratic* trends in these time series:

```{r}
#| code-fold: false

set.seed(123)
funtimes::sync_test(D[, c("Temp_CMIP5", "Temp_obs")] ~ poly(t, 2))
```

Note these results differ substantially based on the window used for computing the WAVK statistic.
The function `funtimes::sync_test()` automatically selects the optimal window based on the heuristic approach of comparing distances between bootstrap distributions [@Lyubchich:Gel:2016:synchronism;@Lyubchich:2016:trends].
:::
