---
output: html_document
editor_options: 
    chunk_output_type: console
---

# Software {.unnumbered}

Examples were calculated in `r R.Version()$version.string` with an effort to use the most recent versions of R packages.

The codes load silently only a few packages:

```{r}
#| eval: false
#| echo: true

library(dplyr)
library(ggplot2)
library(patchwork)
```

All other packages are named before the function as in `forecast::ggAcf()` (this code calls the function `ggAcf()` from the package `forecast`) or called immediately before the necessary function use:

```{r}
#| eval: false
#| echo: true

library(fable)
m <- as_tsibble(Y) %>%
    model(ARIMA(Y ~ 1, ic = "bic"))
report(m)
```

The R packages used in this book include (in alphabetic order):

```{r}
#| echo: false

packages2cite <- sort(c(
    # 'car'
    'dplyr'
    ,'dynlm'
    ,'Ecdat'
    # ,'ecm'
    ,'fable'
    ,'feasts'
    ,'fGarch'
    ,'FinTS'
    ,'fma'
    ,'forecast'
    ,'funtimes'
    ,'gamlss'
    ,'gamlss.util'
    ,'GGally'
    # # ,'ggfortify'
    ,'ggplot2'
    ,'ggpubr'
    ,'Kendall'
    ,'knitr'
    ,'lawstat'
    ,'lmtest'
    ,'lomb'
    # ,'mgcv'
    ,'nlme'
    ,'patchwork'
    ,'plotly'
    ,'pracma'
    ,'randtests'
    ,'readr'
    ,'rmarkdown'
    ,'tseries'
    ,'TSstudio'
    ,'urca'))

# Install the packages, if missing
new.packages <- packages2cite[!(packages2cite %in% installed.packages()[,"Package"])]
if (length(new.packages)) {
    install.packages(new.packages)
}

# Create a bib database for R packages
knitr::write_bib(c(.packages()
                   ,'base'
                   ,packages2cite
), 'refpackages.bib')

# Form a list with citations
tmp <- paste0("* `", packages2cite, "` [@R-", packages2cite, "]")
```

`r paste(tmp, collapse = "\n")`
